A Core Calculus of Metaclasses

### 2004-12-16 metaclasses/semantics-proofs.tex
submitted version of paper

\newcommand{\preservation}{
\begin{theorem}[Subject Reduction]
If \eeproves{\typ{\e}{\SSS}} and \trans{\e}{\ep}
then \eeproves{\typ{\ep}{\T}} where {\proves \emptyset {\subtype{\T}{\SSS}}}.
\end{theorem}
}
\newcommand{\progress}
{
\begin{theorem}[Progress]
If \eeproves{\typ{\e}{\SSS}} then one of the following holds:
\begin{itemize}
\item $\e = \mapping{\of}{\ov \vv}{\GG}$
\item \trans \e \ep
\item $\e = \cast{\SSS}{\ep}$ and $\eeproves{\typ{\ep}{\T}}$ and $\eproves{\notsubtype \T \SSS}$.
\end{itemize}
\end{theorem}
}
\newcommand{\soundness}
{
\begin{theorem}[Soundness]
If \typ{\e}{\SSS} then either
\begin{itemize}
\item $\e \transarrow  \mapping{\of}{\ov \vv}{\GG}$
\item $\e \transarrow \ep$ where $\ep$ is an invalid cast
\item $\e$ reduces infinitely.
\end{itemize}
\end{theorem}
}

@@

Immediate from the premises of \rulename{WF-ClassDef} and \rulename{R-Class}.

@@

We know that $\e = \invoke {\Phi_{\GG}} {\mof \dd}$ and that $\ep =
{\substituteTwoArg \ox \od \this {\Phi_{\GG}}} \e_0$. Further, $\e$
was typed by \rulename{T-Invk} to have type $\U$ where $\mtype \m \GG
= \arrType \oT \U$ and $\eeproves {\typ \od \oTp}$ and $\eproves {\subtype \oTp
\oT}$.  As a premise of \rulename{R-Invk}, we know that $\mbody \m \GG
= (\ox, \e_0')$.

By the lemma $mtype$ and $mbody$ agree, $\proves
{\emptyset;{\typ{\ox}{\oT}},{\typ{\this}{\GG}}} {\typ{\e_0'}{\Up}}$
where $\eproves {\subtype \Up \U}$.  Then by the lemma Substitution
Preserves Typing, $\eeproves {\typ {\ep = {\substituteTwoArg \ox \od
\this {\Phi_{\GG}}} \e_0} \Upp}$ where $\eproves {\subtype \Upp \Up}$.

@@

\proofcase{\rulename{T-Class}} In this case $\e = \CSp$ where $\CSp$ is ground.  Then by the lemma  Agreement of {\it field-vals} and {\it fields}, $\typeparams{\CSp} = \oX \implSym \parens {\oI, \oJ}$ for some     Further, $\fieldVals{\typeof \CSp}  = \Tfe$ for some $\oT$, $\of$, and $\oe$.  Therefore, \rulename{R-Class} applies and $\trans \SSS {\mapping \of {\SpforX \CSp} {\typeof \CSp}}$.

@@

\proofcase{\rulename{T-Invk}}  Here there are two cases.
\begin{itemize}
\item $\e = \invoke \rr {\mof \dd}$ where $\rr$ is not a mapping.  Then by the induction hypothesis, either $\rr$ contains a bad cast or $\trans \rr \rr'$ and \rulename{C-Rcvr} applies.
\item $\e = \invoke {\Phi_{\T}} {\mof \dd}$  We know from the antecedent that $\mtype \m {\bound \T} = \arrType \oU \V$ and therefore $\mtype \m \T = \arrType \oU \V$ since $\T$ is ground.  Therefore, since {\it mbody} is defined everywhere {\it mtype} is defined, $\mbody \m \T = (\ox, \e_0)$ for some $\ox$ and $\e_0$.  Thus \rulename{R-Invk} applies.
\end{itemize}

\proofcase{\rulename{T-Field}} Here there are two cases, either the reciever is a mapping or not.  In the first, by the antecedent of the typing rule, we can lookup the field successfully and apply \rulename{R-Field}.  Otherwise, we can apply \rulename{C-Field}.

@@

\newcommand{\substitutionPreservesTyping}{
\begin{lemma}[Substitution Preserves Typing]
If $\dgprove {\typ \e \T}$ and $\Gamma = \typ{\ox}{\oS}$ and
${\proves {\Delta;\emptyset} {\typ {\od} {\oU}}}$ and ${\dproves {\subtype \oU \oS}}$
then ${\proves {\Delta;\emptyset} {\typ {{\substitute \ox \od} \e} \Tp}}$ where ${\dproves {\subtype{\Tp}{\T}}}$.
\end{lemma}
}

@@

\newcommand{\methodTypingLemma}
{
\begin{lemma}[Subtyping Preserves Method Typing]
If $\mtype \m \U = \arrType \oT \SSS$ and $\proves \emptyset {\subtype \V \U}$ then
$\mtype \m \V = \arrType \oT \SSS$.
\end{lemma}
}

@@

In this case, $\V = \CR$, where $CT(\C) = {\typeFront {\typeArgsBounds \C \X \I \J} \B \A}
\braces{... \oM}$ and $\U = \RforX \A$.  We need to show that $\mtype
\m \CR = \arrType \oT \SSS$.   By lemma $methods$ is well-defined, it
suffices to show that $\simpleMethod \in \methods \CR$.

We know that $\simpleMethod \in \methods \U$.  So, we proceed by
induction on the derivation of $\methods \U$.

@@

\subcase{$\U = \typeArgs \D \W$}
Then, $\U = \RforX \A$.  By \rulename{MethodsClass}, $\methods \V =
\methods \CR = (\RforX \oM)\ \cup\ \methods {\RforX \A} = (\RforX
\oM)\ \cup\ \methods \U$.  Therefore, any method in  $\methods \U$
must be in $\methods \V$.

@@

Analagous to \rulename{S-Super}.

@@

\newcommand{\fieldsPreservedLemma}
{
\begin{lemma}[Fields are Preserved by Subtypes]
If $\fields \U = \oF$ and ${\proves {\emptyset} {\subtype \V \U}}$ then $\exists\ \oG$ where $\fields \V = \oF\ @\ \oG$.
\end{lemma}
}

\newcommand{\fieldsPreservedProof}
{

\myProof
We prove this by induction over the derivation for $\fields \V$.

@@

\proofcase{$\V = \CR$}

@@

Let $\T$ be the intermediate type. Then $\fields \V = {\fields \T}\ @\
{\ov {\HH'}}$ and $\fields \V = {\fields \T}\ @\ {\ov {\HH''}}$ by the
induction hypothesis.  Thus $\oG = {\ov {\HH'}}\ @\ {\ov {\HH''}}$.

\proofcase{$\V = \typeof{\CS}$}

@@

\subcase{\rulename{S-Trans}}
As above.

\subcase{\rulename {S-Kind}}

$\SforX \A = \U$ Then $\fields \V = \SforX \oW\ \oh$.  By
well-formedness of $\C$, we know that $\fields \B \subseteq \oW\ \oh$.
Then $\SforX \fields \B \subseteq \SforX \oW\ \oh$, and $\fields {\SforX \B} \subseteq
\SforX \oW\ \oh$ by lemma Substitution Distributes over $fields$.
\end{description}

@@

\newcommand{\fieldsAgreeLemma}{
\begin{lemma}[Agreement of $fields$ and {\it field-vals}]
If $\fields \GG = \Tf$ and $\fieldVals \GG = \oTp\ \ofp\ \oe$ then
$\oT = \oTp$, $\of = \ofp$ and $\eept \oe \oS$ where $\esub \oS \oT$.
\end{lemma}
}
\newcommand{\fieldsAgreeProof}{
\myProof The proof is exactly analagous to the proof of lemma {\it mtype} and {\it mbody} Agree.
}

@@

\proofcase{\rulename{S-Kind}} Analagous to \rulename{S-Super}.

>>>>

\newcommand{\preservation}{
\begin{theorem}[Subject Reduction]
\mbox{}\\
If \eeproves{\typ{\e}{\SSS}} and \trans{\e}{\ep}
then \eeproves{\typ{\ep}{\T}} where {\proves \emptyset {\subtype{\T}{\SSS}}}.
\end{theorem}
}
\newcommand{\progress}
{
\begin{theorem}[Progress]
\mbox{}\\
If \eeproves{\typ{\e}{\SSS}} then one of the following holds:
\begin{itemize}
\item $\e = \mapping{\of}{\ov \vv}{\GG}$
\item \trans \e \ep
\item $\e = \cast{\SSS}{\ep}$ and $\eeproves{\typ{\ep}{\T}}$ and $\eproves{\notsubtype \T \SSS}$.
\end{itemize}
\end{theorem}
}
\newcommand{\soundness}
{
\begin{theorem}[Soundness]
If \typ{\e}{\SSS} then either
\begin{itemize}
\item $\e \transarrow  \mapping{\of}{\ov \vv}{\GG}$
\item $\e \transarrow \ep$ where $\ep$ is an invalid cast
\item $\e$ reduces infinitely.
\end{itemize}
\end{theorem}
}

@@

Immediate from the premises of \\\rulename{WF-ClassDef} and \rulename{R-Class}.

@@

We know that $\e = \invoke {\Phi_{\GG}} {\mof \dd}$ and that \\$\ep =
{\substituteTwoArg \ox \od \this {\Phi_{\GG}}} \e_0$. Further, $\e$
was typed by \rulename{T-Invk} to have type $\U$ where $\mtype \m \GG
= \arrType \oT \U$ and $\eeproves {\typ \od \oTp}$ and $\eproves {\subtype \oTp
\oT}$.  As a premise of \rulename{R-Invk}, we know that $\mbody \m \GG
= (\ox, \e_0')$.

By the lemma $mtype$ and $mbody$ agree, $\emptyset;{\typ{\ox}{\oT}},$\\
$\proves {{\typ{\this}{\GG}}} {\typ{\e_0'}{\Up}}$ where $\eproves
{\subtype \Up \U}$.  Then by the lemma Substitution Preserves Typing,
$\eeproves {\typ \ep \Upp}$ where $\ep = {\substituteTwoArg \ox \od
    \this {\Phi_{\GG}}} \e_0$ and $\eproves {\subtype \Upp \Up}$.

@@

\proofcase{\rulename{T-Class}} In this case $\e = \CSp$ where $\CSp$
is ground.  Then by the lemma Agreement of {\it field-vals} and {\it
  fields}, $\fieldVals{\typeof \CSp} = \Tfe$ for some $\oT$, $\of$,
and $\oe$.  Further, $\typeparams{\CSp} = \oX \superSym \oI \kindSym
  \oJ$ for some $\oX, \oI, \oJ$. Therefore, \rulename{R-Class}
applies and \\$\trans \SSS {\mapping \of {\SpforX \CSp} {\typeof
    \CSp}}$.

@@

\proofcase{\rulename{T-Invk}}  Here there are two cases.
\begin{itemize}
\item $\e = \invoke \rr {\mof \dd}$ where $\rr$ is not a mapping.  Then by the induction hypothesis, either $\rr$ contains a bad cast or $\trans \rr \rr'$ and \rulename{C-Rcvr} applies.
\item $\e = \invoke {\Phi_{\T}} {\mof \dd}$  We know from the antecedent that \\$\mtype \m {\bound \T} = \arrType \oU \V$ and therefore $\mtype \m \T = \arrType \oU \V$ since $\T$ is ground.  Therefore, since {\it mbody} is defined everywhere {\it mtype} is defined, $\mbody \m \T = (\ox, \e_0)$ for some $\ox$ and $\e_0$.  Thus \rulename{R-Invk} applies.
\end{itemize}

\proofcase{\rulename{T-Field}} Here there are two cases, either the receiver is a mapping or not.  In the first, by the antecedent of the typing rule, we can lookup the field successfully and apply \rulename{R-Field}.  Otherwise, we can apply \rulename{C-Field}.

@@

\newcommand{\substitutionPreservesTyping}{
\begin{lemma}[Substitution Preserves Typing]
\mbox{}\\
If $\dgprove {\typ \e \T}$ and $\Gamma = \typ{\ox}{\oS}$ and
${\proves {\Delta;\emptyset} {\typ {\od} {\oU}}}$ and ${\dproves {\subtype \oU \oS}}$
then ${\proves {\Delta;\emptyset} {\typ {{\substitute \ox \od} \e} \Tp}}$ where ${\dproves {\subtype{\Tp}{\T}}}$.
\end{lemma}
}

@@

\newcommand{\methodTypingLemma}
{
\begin{lemma}[Subtyping Preserves Method Typing]
\mbox{}\\
If $\mtype \m \U = \arrType \oT \SSS$ and $\proves \emptyset {\subtype \V \U}$ then
$\mtype \m \V = \arrType \oT \SSS$.
\end{lemma}
}

@@

In this case, $\V = \CR$, where \\$CT(\C) = {\typeFront {\typeArgsBounds \C \X \I \J} \B \A}
\braces{... \oM}$ \\and $\U = \RforX \A$.  We need to show that $\mtype
\m \CR = \arrType \oT \SSS$.   By lemma $methods$ is well-defined, it
suffices to show that $\simpleMethod \in \methods \CR$.

Since we know that $\simpleMethod \in$ \\$\methods \U$, we proceed by
induction on the derivation of $\methods \U$.

@@

\subcasenl{$\U = \typeArgs \D \W$}\\
Then, $\U = \RforX \A$.  By \rulename{MethodsClass}, $\methods \V =
\methods \CR = (\RforX \oM)\ \cup\ \methods {\RforX \A} = (\RforX
\oM)\ \cup\ \methods \U$.  Therefore, any method in  $\methods \U$
must be in $\methods \V$.

@@

Analogous to \rulename{S-Super}.

@@

\newcommand{\fieldsPreservedLemma}
{
\begin{lemma}[Fields are Preserved by Subtypes]
\mbox{}\\
If $\fields \U = \oF$ and ${\proves {\emptyset} {\subtype \V \U}}$ then $\exists\ \oG$ where $\fields \V = \oF\ @\ \oG$.
\end{lemma}
}

\newcommand{\fieldsPreservedProof}
{
\myProof
We prove this by induction over the derivation for $\fields \V$.

@@

\proofcasenl{$\V = \CR$}

@@

Let $\T$ be the intermediate type. Then \\$\fields \V = {\fields \T}\ @\
{\ov {\HH'}}$ and $\fields \V = {\fields \T}\ @\ {\ov {\HH''}}$ by the
induction hypothesis.  Thus $\oG = {\ov {\HH'}}\ @\ {\ov {\HH''}}$.

\proofcasenl{$\V = \typeof{\CS}$}

@@

\subcase{\rulename{S-Trans}}
As above.
\newpage
\subcasenl{\rulename {S-Kind}}

$\SforX \A = \U$ Then $\fields \V = \SforX \oW\ \oh$.  By
well-formedness of $\C$, we know that $\fields \B \subseteq \oW\ \oh$.
Then $\SforX \fields \B \subseteq \SforX \oW\ \oh$, and $\fields
{\SforX \B} \subseteq$\\$\SforX \oW\ \oh$ by lemma Substitution
Distributes over $fields$.
\end{description}

@@

\newcommand{\fieldsAgreeLemma}{
\begin{lemma}[Agreement of $fields$ and {\it field-vals}]
\mbox{}\\
If $\fields \GG = \Tf$ and $\fieldVals \GG = \oTp\ \ofp\ \oe$ then
$\oT = \oTp$, $\of = \ofp$ and $\eept \oe \oS$ where $\esub \oS \oT$.
\end{lemma}
}
\newcommand{\fieldsAgreeProof}{
\myProof The proof is exactly analogous to the proof of lemma {\it mtype} and {\it mbody} Agree.
}

@@

\proofcase{\rulename{S-Kind}} Analogous to \rulename{S-Super}.
